Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2023-10-03 paper/paper-infer.tex
some formating

Most traits and diseases in humans are heritable.
What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero effect on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
As for the distribution of causal effects relative to their allele frequencies, it is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
In this model, the expected phenotypic variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$. 
These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher} ($h^2$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).

@@

@@

The HapMap3 variants generally provide a good coverage of the whole genome.
We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ `A' and `T', or `C' and `G'), and which includes SNPs only (e.g.\ no indel).
Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
There are around 11.5M such variants.
Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
Finally, we design an algorithm which aims at maximizing the tagging of all these variants read. We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set, which we call HapMap3+.
We start by including all previously used HapMap3 variants.
Then, for the sake of simplicity, we use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
Note that we only allow non-ambiguous SNPs to be included.
This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \citet{prive2021optimal} to obtain 431 independent LD blocks.
Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (resp.\ 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.8\textwidth]{ukbb_r2_MLE}}
	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank. These are stratified by the polygenicity estimated from LDpred2-auto. Green dashed lines represent the 1:1 line. The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $r^2$ estimates from the iterations (after burn-in) of the chains kept. The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap. Colors represent the number of chains kept (out of 50). ``F\_height'' and ``M\_height'' use females and males only, respectively (in both GWAS and test sets). \label{fig:ukbb_r2}}
\end{figure}

@@

In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure S36.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one LD block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure S37).

@@

\section*{Funding}

@@

The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
Descriptions of UK Biobank phenotypes used here are available at \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}.
Simulation and UKBB results are available at \url{https://github.com/privefl/paper-infer/tree/main/results}.
LD matrices for HapMap3+ variants computed from the N.W. European UKBB data used in this paper are available at \url{https://doi.org/10.6084/m9.figshare.21305061}.
All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/main/code}.
We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN.

>>>>

Most traits and diseases in humans are heritable.
What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero effect on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
As for the distribution of causal effects relative to their allele frequencies, it is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
In this model, the expected phenotypic variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$. 
These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).

@@

We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.

@@

The HapMap3 variants generally provide a good coverage of the whole genome.
We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ `A' and `T', or `C' and `G'), and which includes SNPs only (e.g.\ no indel).
Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
There are around 11.5M such variants.
Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
Finally, we design an algorithm which aims at maximizing the tagging of all these variants read. We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set, which we call HapMap3+.
We start by including all previously used HapMap3 variants.
Then, for the sake of simplicity, we use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
Note that we only allow non-ambiguous SNPs to be included.
This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \citet{prive2021optimal} to obtain 431 independent LD blocks.
Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (respectively 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.8\textwidth]{ukbb_r2_MLE}}
	\caption{Inferred predictive performance $r^2$ in the UK Biobank. This compares the $r^2$ inferred from the Gibbs sampler of LDpred2-auto (Methods) versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank. These are stratified by the polygenicity estimated from LDpred2-auto. Green dashed lines represent the 1:1 line. The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $r^2$ estimates from the iterations (after burn-in) of the chains kept. The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap. Colors represent the number of chains kept (out of 50). ``F\_height'' and ``M\_height'' use females and males only, respectively (in both GWAS and test sets). \label{fig:ukbb_r2}}
\end{figure}

@@

In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure S36.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one LD block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (respectively 8) of them have at least 50\% (respectively 80\%) of their heritability explained by a single block (Figure S37).

@@

@@

The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
UK Biobank received ethical approval from the NHS National Research Ethics Service North West (11/NW/0382). The present analyses were conducted under UK Biobank data application number 58024.
All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/main/code}.

\section*{Web Resources}

\noindent Descriptions of UK Biobank phenotypes used here, \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}

\noindent Simulation and UKBB results from this study,  \url{https://github.com/privefl/paper-infer/tree/main/results}

\noindent LD matrices for HapMap3+ variants computed from the N.W. European UKBB data used in this paper, \url{https://doi.org/10.6084/m9.figshare.21305061}

\noindent A tutorial on running LDpred2 with R package bigsnpr (for both prediction and inference), \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}
